---
title: "Assignment 1: Student Spending Simulation and Analysis"
author: "Group Analysis"
date: "`r Sys.Date()`"
output: word_document
---

```{r variables}
# This hidden chunk sets global options to ensure code and output are both visible
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

set.seed(2026)
n <- 4860

# 1. Independent Variables
x_gender <- sample(c(0, 1), n, replace = TRUE) 
x_funding <- sample(c(0, 1), n, replace = TRUE, prob = c(0.4, 0.6))
x_accommodation <- sample(c(1, 2, 3), n, replace = TRUE, prob = c(0.3, 0.5, 0.2))

# Distance: Continuous KM
x_distance <- ifelse(x_accommodation == 1, runif(n, 0.01, 0.49), runif(n, 1.5, 15))
x_distance <- round(x_distance, 2)

# Financial Literacy & Study Hours
x_fin_lit <- round(pmax(pmin(rnorm(n, 55, 12), 100), 0), 2)
x_study_hrs <- round(runif(n, 9, 42) * 4, 2)

# Social Outings (Correlation: Literacy and Study Hours reduce outings)
lambda_outings <- pmax(1, 14 - (0.05 * x_study_hrs) - (0.06 * x_fin_lit))
x_outings <- rpois(n, lambda = lambda_outings)
```


ghhj
 
```{r model}
# 2. Generating Spending (Y) with high noise for 35-50% R-squared
beta_0 <- 2200 
acc_adj <- ifelse(x_accommodation == 2, 850, ifelse(x_accommodation == 3, 3200, 0))
# Large SD (1650) ensures the model doesn't explain too much of the variance
epsilon <- rnorm(n, mean = 0, sd = 1650) 
# Base spending (Intercept)
beta_0 <- 2200 

# Individual effects
 # Baseline spending
beta_funding  <- 500   # NSFAS increase
beta_dist     <- 25    # Cost per KM
beta_lit      <- -12   # Savings per literacy point
beta_outings  <- 180   # Cost per social outing
beta_off_campus <- 850
beta_apartment  <- 3200

# High noise to ensure R-squared is 35-50%
sigma_noise   <- 1650

y_spending <- beta_0 + (beta_funding * x_funding) + acc_adj + (beta_dist * x_distance) + 
              (beta_outings * x_outings) + (beta_lit * x_fin_lit) + epsilon

# Rounding and realistic constraints (R1200 - R13000)
y_spending <- round(pmax(pmin(y_spending, 13000), 1200), 2)

obs_data <- data.frame(Spending = y_spending, Funding = x_funding, 
                       Accommodation = x_accommodation, Distance = x_distance, 
                       FinLit = x_fin_lit, StudyHrs = x_study_hrs, 
                       Outings = x_outings, Gender = x_gender)

```
ggggggggggg  
```{r histograms}
# --- HISTOGRAMS FOR OBSERVATIONAL DATA ---

# Set up a grid of 3 rows and 3 columns
par(mfrow = c(3, 3), mar = c(4, 4, 2, 1)) 

# 1. Spending (The Outcome)
hist(obs_data$Spending, col = "skyblue", main = "Spending (Y)", xlab = "Rands")

# 2. Funding
hist(obs_data$Funding, col = "lightgreen", main = "Funding Status", xlab = "0 = No, 1 = Yes")

# 3. Accommodation
hist(obs_data$Accommodation, col = "salmon", main = "Accommodation Type", xlab = "1=Res, 2=Off, 3=Apt")

# 4. Distance
hist(obs_data$Distance, col = "gold", main = "Distance to Campus", xlab = "Kilometers")

# 5. Financial Literacy
hist(obs_data$FinLit, col = "plum", main = "Fin. Literacy Score", xlab = "Score (0-100)")

# 6. Study Hours
hist(obs_data$StudyHrs, col = "lightblue", main = "Study Hours", xlab = "Weekly Hours")

# 7. Social Outings
hist(obs_data$Outings, col = "orange", main = "Social Outings", xlab = "Number of Outings")

# 8. Gender (Nuisance)
hist(obs_data$Gender, col = "grey", main = "Gender", xlab = "0 = Male, 1 = Female")

# Reset plotting parameters to default (1x1)
par(mfrow = c(1, 1))
```
gggg
```{r scatter plot mstrix}

# --- SCATTER PLOT MATRIX (OBSERVATIONAL DATA) ---

# Select the key variables for the matrix
# Spending, Distance, FinLit, StudyHrs, and Outings
plot_data <- obs_data[, c("Spending", "Distance", "FinLit", "StudyHrs", "Outings")]

# Generate the matrix
# pch=16 and col=rgb(...) makes the points small and transparent to handle 4860 rows
pairs(plot_data, 
      main = "Scatter Plot Matrix: Student Spending Drivers",
      pch = 16, 
      cex = 0.5, 
      col = rgb(0.2, 0.4, 0.6, 0.1),
      lower.panel = panel.smooth) # Adds a trend line to the bottom half
```
tttt

```{r plot}
hist(obs_data$Spending, main="Observational: Monthly Student Spending", 
     xlab="ZAR", col="skyblue", border="white", breaks=40)
abline(v = c(2500, 5000), col = "red", lwd = 2, lty = 2)

# Relationship between Outings and Spending
plot(obs_data$Outings, obs_data$Spending, main="Outings vs Spending",
     xlab="Number of Outings", ylab="Spending (ZAR)", pch=16, col=rgb(0,0,1,0.2))


```
tttt

```{r textiles}
# Create Balanced Grid
exp_grid <- expand.grid(Funding = c(0, 1), 
                        Accommodation = c(1, 2, 3), 
                        Distance = c("Close", "Far"), 
                        FinLit = c("Low", "Moderate", "High"), 
                        StudyHrs = c("Low", "Moderate", "High"), 
                        Outings = c("Low", "Moderate", "High"))

# Replicate to reach N=4860
exp_data <- exp_grid[rep(seq_len(nrow(exp_grid)), each = 15), ]

# Re-simulate Response using categorical effects
dist_eff <- ifelse(exp_data$Distance == "Far", 700, 0)
lit_eff  <- ifelse(exp_data$FinLit == "Moderate", -300, 
                   ifelse(exp_data$FinLit == "High", -800, 0))
out_eff  <- ifelse(exp_data$Outings == "Moderate", 1000, 
                   ifelse(exp_data$Outings == "High", 2400, 0))
acc_eff  <- ifelse(exp_data$Accommodation == 2, 850, 
                   ifelse(exp_data$Accommodation == 3, 3200, 0))

# Apply the same noise level to keep R-squared around 35-50%
exp_data$Spending <- round(beta_0 + (500 * exp_data$Funding) + acc_eff + 
                           dist_eff + lit_eff + out_eff + 
                           rnorm(nrow(exp_data), 0, 1650), 2)

exp_data$Spending <- pmax(pmin(exp_data$Spending, 13000), 1200)

model_obs <- lm(Spending ~ Funding + factor(Accommodation) + Distance + 
                FinLit + StudyHrs + Outings + Gender, data = obs_data)

```
jjjjjj

```{r R-squared}
# Show coefficients and R-squared
round(summary(model_obs)$coefficients, 2)
cat("Observational R-Squared:", round(summary(model_obs)$r.squared, 2))

model_exp <- lm(Spending ~ Funding + factor(Accommodation) + Distance + 
                FinLit + StudyHrs + Outings, data = exp_data)

# Show coefficients and R-squared
round(summary(model_exp)$coefficients, 2)
cat("Experimental R-Squared:", round(summary(model_exp)$r.squared, 2))
par(mfrow = c(2, 2))
plot(model_exp)

```

## Comparison of True vs. Estimated Parameters (Experimental Data)

This section evaluates the accuracy of the Experimental Model. We compare the coefficients estimated by the model against the discrete categorical effects we programmed into the simulation.

```{r exp_comparison_table_expda}
# 1. Extract Estimated Coefficients from the Experimental Model
exp_estimates <- round(coef(model_exp), 2)

# 2. Define the True Betas for the Experimental Categories
# These match the 'dist_eff', 'lit_eff', and 'out_eff' values from Part III
exp_true_values <- c(
  "(Intercept)"             = beta_0,
  "Funding"                 = beta_funding,
  "factor(Accommodation)2"  = beta_off_campus,
  "factor(Accommodation)3"  = beta_apartment,
  "DistanceFar"             = 700,   # Programmed effect for Far vs Close
  "FinLitModerate"          = -300,  # Programmed effect for Moderate vs Low
  "FinLitHigh"              = -800,  # Programmed effect for High vs Low
  "StudyHrsModerate"        = 0,     # Programmed as 0 (indirect effect only)
  "StudyHrsHigh"            = 0,     # Programmed as 0 (indirect effect only)
  "OutingsModerate"         = 1000,  # Programmed effect for Moderate vs Low
  "OutingsHigh"             = 2400   # Programmed effect for High vs Low
)

# 3. Align and combine into a comparison data frame
# This ensures we only compare variables that exist in the model output
exp_comparison <- data.frame(
  Variable = names(exp_estimates),
  True_Beta = exp_true_values[names(exp_estimates)],
  Estimated_Beta = exp_estimates,
  Bias = round(exp_estimates - exp_true_values[names(exp_estimates)], 2)
)

# 4. Display the table in the Word Document
knitr::kable(exp_comparison, 
             caption = "Experimental Model: Programmed Effects vs. Model Estimates",
             row.names = FALSE)
```
hhhh  
j
```{r Comparision_y}
# --- COMPARATIVE EXPENDITURE HISTOGRAMS ---

# Set layout for 2 plots side-by-side
par(mfrow = c(1, 2), mar = c(5, 4, 4, 2))

# 1. Observational Spending Distribution
hist(obs_data$Spending, 
     main = "Observed Spending", 
     xlab = "Monthly Expenditure (ZAR)", 
     col = "#5DADE2", # Soft Blue
     border = "white",
     breaks = 40,
     xlim = c(1200, 13000))
# Add reference lines for the R2.5k - R5k target range
abline(v = c(2500, 5000), col = "darkred", lwd = 2, lty = 2)

# 2. Experimental Spending Distribution
hist(exp_data$Spending, 
     main = "Experimental Spending", 
     xlab = "Monthly Expenditure (ZAR)", 
     col = "#58D68D", # Soft Green
     border = "white",
     breaks = 40,
     xlim = c(1200, 13000))
# Add same reference lines
abline(v = c(2500, 5000), col = "darkred", lwd = 2, lty = 2)

# Reset layout
par(mfrow = c(1, 1))
```
Feature,Observational Plot,Experimental Plot
Shape,"Likely more ""Bell-Shaped"" (Normal) because continuous variables average out.","Often looks more ""Blocky"" or Uniform because we have equal numbers of students in every category."
Range,"Both should strictly stay between R1,200 and R13,000.","Both should strictly stay between R1,200 and R13,000."
Peak (Mode),"Should center around the R2,500 – R5,000 mark (indicated by red lines).","Should also center around the same mark, verifying the simulation logic is identical."





## Distributional Comparison: Superimposed Histogram and Q-Q Plots

This section visually compares the "Expenditure" (Spending) between the Observational and Experimental groups to verify that both simulations reflect the same South African student financial context.

```{r distribution_plots}
# 1. SETUP: Superimposed Histogram
# Calculate common range for the x-axis
xlim_range <- range(c(obs_data$Spending, exp_data$Spending))

# Create the first histogram (Observational)
hist(obs_data$Spending, 
     breaks = 40, 
     col = rgb(0.2, 0.5, 0.8, 0.5), # Transparent Blue
     xlab = "Monthly Expenditure (ZAR)", 
     main = "Superimposed Expenditure Distribution",
     xlim = xlim_range,
     border = "white")

# Superimpose the second histogram (Experimental)
hist(exp_data$Spending, 
     breaks = 40, 
     col = rgb(0.8, 0.2, 0.2, 0.5), # Transparent Red
     add = TRUE, 
     border = "white")

# Add a legend to distinguish the two
legend("topright", 
       legend = c("Observational", "Experimental"), 
       fill = c(rgb(0.2, 0.5, 0.8, 0.5), rgb(0.8, 0.2, 0.2, 0.5)),
       border = "black")

# 2. SETUP: Side-by-Side Q-Q Plots
# These check if the spending follows a normal distribution
par(mfrow = c(1, 2)) # 1 row, 2 columns

# Q-Q Plot for Observational
qqnorm(obs_data$Spending, 
       main = "Q-Q Plot: Observational", 
       col = rgb(0.2, 0.5, 0.8, 0.4), 
       pch = 16)
qqline(obs_data$Spending, col = "darkblue", lwd = 2)

# Q-Q Plot for Experimental
qqnorm(exp_data$Spending, 
       main = "Q-Q Plot: Experimental", 
       col = rgb(0.8, 0.2, 0.2, 0.4), 
       pch = 16)
qqline(exp_data$Spending, col = "darkred", lwd = 2)

# Reset layout
par(mfrow = c(1, 1))
```
## Model Performance Comparison: R-Squared and Residuals

This table evaluates the overall "Success" of our two models. While we targeted an R-squared of 35-50%, the residuals provide a deeper look into the prediction error of each design.

```{r model_fit_comparison}
# 1. Extract R-Squared and Adjusted R-Squared
r2_obs <- summary(model_obs)$r.squared
adj_r2_obs <- summary(model_obs)$adj.r.squared

r2_exp <- summary(model_exp)$r.squared
adj_r2_exp <- summary(model_exp)$adj.r.squared

# 2. Extract Residual Statistics (RSE and Range)
# Residual Standard Error (RSE) represents the "Average Error" in Rands
rse_obs <- summary(model_obs)$sigma
rse_exp <- summary(model_exp)$sigma

# Range of Residuals (Min to Max)
res_range_obs <- range(residuals(model_obs))
res_range_exp <- range(residuals(model_exp))

# 3. Construct the Comparison Table
model_perf_table <- data.frame(
  Metric = c("R-Squared (Raw)", "Adjusted R-Squared", "Residual Std. Error (RSE)", "Min Residual", "Max Residual"),
  Observational_Model = c(round(r2_obs, 4), round(adj_r2_obs, 4), round(rse_obs, 2), 
                          round(res_range_obs[1], 2), round(res_range_obs[2], 2)),
  Experimental_Model = c(round(r2_exp, 4), round(adj_r2_exp, 4), round(rse_exp, 2), 
                         round(res_range_exp[1], 2), round(res_range_exp[2], 2))
)

# 4. Display the table
knitr::kable(model_perf_table, 
             caption = "Performance Summary: Observational vs. Experimental Models",
             col.names = c("Performance Metric", "Observational Data", "Experimental Data"))
```
```{r savinh}
# =========================================================
# PART VII: SAVING THE SIMULATED DATASETS
# =========================================================

# 1. Save the Observational Data
# This file contains the continuous variables and the nuisance variable (Gender)
write.csv(obs_data, "Student_Spending_Observational.csv", row.names = FALSE)

# 2. Save the Experimental Data
# This file contains the categorized/balanced variables used for the True Model check
write.csv(exp_data, "Student_Spending_Experimental.csv", row.names = FALSE)

# Confirmation message to be printed in the document
cat("Success! The following files have been saved to your working directory:\n",
    "- Student_Spending_Observational.csv\n",
    "- Student_Spending_Experimental.csv")
```